---
title: "Inference for SIR models"
author: "B. Raynor"
date: "7/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## References 
This code uses the MultiBD package - code example adapted from package [vignette](Tutorial source: https://cran.r-project.org/web/packages/MultiBD/vignettes/SIR-MCMC.pdf)

The inference in this code relies on the Metroplis Hasting sampler. This is one of the work horse MCMC samplers used in Bayesian statistics. There are a ton of resources and lectures online that detail MH samplers if you want to learn more about the statistics.  

+ [Very cited paper](https://eml.berkeley.edu/reprints/misc/understanding.pdf)
+ [Tutorial w/ R code](https://www.aptech.com/resources/tutorials/bayesian-fundamentals/metropolis-hastings-sampler/)
+ [Youtube video tutorial](https://www.youtube.com/watch?v=yCv2N7wGDCw&ab_channel=ritvikmath)

## Code

```{r, warning=FALSE, message=FALSE}
#clear environment
rm(list = ls()) 

#install.packages("MultiBD")
#install.packages("MCMCpack", repos = 'http://cran.us.r-project.org')
#install.packages("BiocManager")

#load packages
library(MultiBD)
library(BiocManager)
library(MCMCpack)
library(dplyr)
library(ggplot2)
library(deSolve)
```


First, we will simulate data that we know the true parameter values of. Here is a simple SIR model with $\beta = 0.02$ and $\alpha = 2$ with some random noise added.  
```{r, warning=FALSE, message=FALSE}
#Set initial conditions
init <- c(S=200, I=5, R=0)  # popn initial values
parameters <- c(b= 0.02,
                a=2)

#set time series
times <- seq(0, 4, by = 0.5) #unit=days, seq(start day, end day, step)

#Model function
disease_dynamics <- function(time, state, parameters) {
  with(
    as.list(c(state, parameters)), {
      #Equations
      dS <- - b*S*I
      dI <- b*S*I - a*I 
      dR <- a*I
      
      return(list(c(dS, dI, dR)))
    }
  )
}

set.seed(123)
df <-  as.data.frame(ode(y = init, times = times, func = disease_dynamics, parms = parameters))%>%
  mutate(S= round(S + rnorm(length(times), 0,1)),
         I= round(I + rnorm(length(times), 0,1)),
         R= round(R + rnorm(length(times), 0,1)))%>%
  mutate(S=ifelse(S < 0, 0, S),
         I= ifelse(I < 0, 0, I),
         R= ifelse(R < 0, 0, R)
         )
```

Our resulting data that we will now try to fit parameter values is:  
```{r, warning=FALSe, message=FALSE}
knitr::kable(df)
```

Visualized:  
```{r, warning=FALSe, message=FALSE}
#Create color palette
Pal1 = c("S" = "goldenrod", 
         "I" = "red3",
         "R" = "dodgerblue3")

#plot
fig <- ggplot() +
  theme_classic()+
  geom_point(data=df, aes(x=time, y=S, color="S"), size=3)+
  geom_point(data=df, aes(x=time, y=I, color="I"), size=3)+
  geom_point(data=df, aes(x=time, y=R, color="R"), size=3)+
  geom_line(data=df, aes(x=time, y=S, color="S"), size=1, linetype = "dotted")+
  geom_line(data=df, aes(x=time, y=I, color="I"),linetype = "dotted", size=1)+
  geom_line(data=df, aes(x=time, y=R, color="R"),linetype = "dotted", size=1)+
  scale_color_manual(values=Pal1, name="Disease state")+
  theme(text = element_text(size = 20))+
  ylab("Number of individuals") + xlab("Time")
fig
```


To fit the 'unknown' parameter values, we will use the metropolis hasting algorithm. 

Specify functions for Metropolis hasting
```{r, warning=FALSE, message=FALSE}
#log liklihood of SIR
loglik_sir <- function(param, data) {
  alpha <- exp(param[1]) # Rates must be non-negative
  beta <- exp(param[2])
  # Set-up SIR model
  drates1 <- function(a, b) { 0 }
  brates2 <- function(a, b) { 0 }
  drates2 <- function(a, b) { alpha * b }
  trans12 <- function(a, b) { beta * a * b }
  sum(sapply(1:(nrow(data) - 1), # Sum across all time steps k
  function(k) {
    log(
    dbd_prob( # Compute the transition probability matrix
    t = data$time[k + 1] - data$time[k], # Time increment
    a0 = data$S[k], b0 = data$I[k], # From: S(t_k), I(t_k)
    drates1, brates2, drates2, trans12,
    a = data$S[k + 1], B = data$S[k] + data$I[k] - data$S[k + 1],
    computeMode = 4, nblocks = 80 # Compute using 4 threads
    )[1, data$I[k + 1] + 1] # To: S(t_(k+1)), I(t_(k+1))
    )
  }))
}

#log priors for alpha and beta
logprior <- function(param) {
  log_alpha <- param[1]
  log_beta <- param[2]
  dnorm(log_alpha, mean = 0, sd = 100, log = TRUE) +
  dnorm(log_beta, mean = 0, sd = 100, log = TRUE)
}
```

Simulate posterior distribution via Metropolis hasting
```{r, message = FALSE, warning=FALSE}
set.seed(1234)

#set initial conditions
alpha0 <- 2
beta0 <- 0.01

#Find posterior sample
post_sample <- MCMCmetrop1R(fun = function(param) { loglik_sir(param, df) + logprior(param) },
                            theta.init = log(c(alpha0, beta0)),
                            mcmc = 1000, burnin = 200)
#format posterior sample
post_sample <- post_sample%>%
  as.data.frame()%>%
  rename(log_alpha=V1, log_beta=V2)%>%
  mutate(sim = 1:nrow(post_sample))
```

What does the posterior sample look like?  
```{r, message=FALSE, warning=FALSE}
#create color palette
Pal2 = c("log alpha" = "goldenrod", 
         "log beta" = "dodgerblue")

Pal3 = c("alpha" = "goldenrod", 
         "beta" = "dodgerblue")

#visualize metropolis hasting steps
fig3 <- ggplot()+
  theme_classic()+
  geom_line(data=post_sample, aes(y=log_alpha, x=sim, color="log alpha"))+
  geom_line(data=post_sample, aes(y=log_beta, x=sim, color="log beta"))+
  scale_color_manual(values=Pal2, name="Parameter")+
  theme(text = element_text(size = 20))+
  ylab("Parameter estimate") + xlab("Simulation number")+
  ggtitle("Metropolis Hasting steps")

#visualize posterior distributions
fig1 <- ggplot()+
  theme_classic()+
  geom_histogram(data=post_sample, aes(x=exp(log_alpha), fill="alpha"), bins=30, color="black")+
  scale_fill_manual(values=Pal3, name="Parameter")+
  theme(text = element_text(size = 20))+
  xlab("alpha posterior samples")
fig2 <- ggplot()+
  theme_classic()+
  geom_histogram(data=post_sample, aes(x=exp(log_beta), fill="beta"), bins=30, color="black")+
  scale_fill_manual(values=Pal3, name="Parameter")+
  theme(text = element_text(size = 20))+
  xlab("beta posterior samples")
ggpubr::ggarrange(fig1, fig2, common.legend = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=12}
fig3
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=12}
ggpubr::ggarrange(fig1, fig2, common.legend = TRUE)
```

Examine fit:
```{r, warning=FALSE, message=FALSE}
#Set initial conditions
init <- c(S=df$S[1], I=df$I[1], R=df$R[1])  # popn initial values
parameters <- c(b= exp(mean(post_sample$log_beta)),
                a=exp(mean(post_sample$log_alpha))
                )

#set time series
times <- seq(0, tail(df$time, n=1), by = tail(df$time, n=1)/nrow(df)) #unit=days, seq(start day, end day, step)

#Model function
disease_dynamics <- function(time, state, parameters) {
  with(
    as.list(c(state, parameters)), {
      #Equations
      dS <- - b*S*I
      dI <- b*S*I - a*I 
      dR <- a*I
      
      return(list(c(dS, dI, dR)))
    }
  )
}

#Run model
df.out <- as.data.frame(ode(y = init, times = times, func = disease_dynamics, parms = parameters))

#compare parameter estimates
fit <- data.frame(parameter = c("beta", "alpha"), 
                  true_value=c(0.02,2),
                  mean_post = c(exp(mean(post_sample$log_beta)), exp(mean(post_sample$log_alpha))),
                  lower_95CI = c(quantile(exp(post_sample$log_beta), probs = c(0.025))[[1]],
                                 quantile(exp(post_sample$log_alpha), probs = c(0.025))[[1]]),
                  upper_95CI = c(quantile(exp(post_sample$log_beta), probs = c(0.975))[[1]],
                                 quantile(exp(post_sample$log_alpha), probs = c(0.975))[[1]])
                  )%>%
  mutate(mean_post = round(mean_post, 2),
         lower_95CI = round(lower_95CI, 2),
         upper_95CI = round(upper_95CI, 2))

knitr::kable(fit)
```

Visualize:
```{r, warning=FALSE, message=FALSE}
#add fit lines to data plot
fig + 
  geom_line(data=df.out, aes(x=times, y=S, color="S"), size=2, alpha=0.5)+ 
  geom_line(data=df.out, aes(x=times, y=I, color="I"), size=2, alpha=0.5)+
  geom_line(data=df.out, aes(x=times, y=R, color="R"), size=2, alpha=0.5)

```