---
title: "Spatial epidemiology"
author: "B. Raynor"
date: "6/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
rm(list = ls()) 
```


### Identifying clustering   

**Definitions:**  

+ Cluster:  an aggregation of cases grouped in place and time that are suspected to be greater than the number expected, even though the expected number may not be known  [CDC](https://www.cdc.gov/csels/dsepd/ss1978/lesson1/section11.html#:~:text=Outbreak%20carries%20the%20same%20definition,number%20may%20not%20be%20known))  

+ Epidemic: an increase, often sudden, in the number of cases of a disease above what is normally expected in that population in that area  [CDC](https://www.cdc.gov/csels/dsepd/ss1978/lesson1/section11.html#:~:text=Outbreak%20carries%20the%20same%20definition,number%20may%20not%20be%20known))  

+ Outbreak: Same as an epidemic, but is often used for a more limited geographic area  [CDC](https://www.cdc.gov/csels/dsepd/ss1978/lesson1/section11.html#:~:text=Outbreak%20carries%20the%20same%20definition,number%20may%20not%20be%20known))  

+ Hot spot: less strictly defined... [Lessler et al](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5462559/) "In infectious disease epidemiology, “hotspot” is frequently used to refer to areas of elevated disease burden or high transmission efficiency." 

#### **scan statistic**  
A statistic used to evaluate spatial/temporal clusters:  
*"The circular spatial scan statistic is defined through a large number of overlapping circles [18]. For each circle z, a log likelihood ratio LLR(z) is calculated, and the test statistic is defined as the maximum LLR over all circles. The scanning window will depend on the application"* -[Kulldorff et al](https://ij-healthgeographics.biomedcentral.com/articles/10.1186/1476-072X-8-58)

+ Free online software Ricardo recommended [SaTScan](https://www.satscan.org/).  

+ [rflexscan R package](https://www.jstatsoft.org/article/view/v099i13)

Code below is adapted from [rflexscan](https://www.jstatsoft.org/article/view/v099i13) R package code and accompanying paper 
```{r, warning=FALSE, message=FALSE}
library("rgdal")
library("spdep")
library("rflexscan")
library("RColorBrewer")

# for replication
set.seed(12345)


## ----load NYS cancer data-----------------------------------------------------
# Download data
temp <- tempfile()
download.file("https://www.satscan.org/datasets/nyscancer/NYS_Cancer.zip", temp)
unzip(temp, exdir = "NYS_Cancer")

nys <- readOGR("NYS_Cancer/NYSCancer_region.shp",
               stringsAsFactors = FALSE, verbose=FALSE)


## ----figure 1-----------------------------------------------------------------
manhattan <- nys[grep("36061", nys$DOHREGION),]
SIR <- manhattan$OBREAST / manhattan$EBREAST
nsteps <- 7
brks <- c(0, 1.0, 1.2, 1.4, 1.6, 1.8, 2, 10)
brks[1] <- 0
cols <- colorRampPalette(c("white","royal blue"))(nsteps)
grps <- as.ordered(cut(SIR, brks, include.lowest = TRUE))
par(mar = c(1, 1, 1, 1), oma = c(1, 1, 1, 1))
plot(manhattan, col=cols[unclass(grps)], lwd = 0.1)
box()
legend("topleft",
  legend = c("> 2.00", "1.80 - 2.00", "1.60 - 1.80",
    "1.40 - 1.60", "1.20 - 1.40", "1.00 - 1.20", "< 1.00"), 
  fill=rev(cols), bty="n") 
```

*"Figure 1: Standardized incidence ratios of breast cancer (female) in the Manhattan borough
of New York City for the years 2005–2009 based on the 2010 census counts."* [Otani and Takahashi, 2001](https://www.jstatsoft.org/article/view/v099i13)

```{r,warning=FALSE, message=FALSE}
## ----extract Manhattan data---------------------------------------------------
manhattan <- nys[startsWith(nys$DOHREGION, "36061"),]


## ----transformation-----------------------------------------------------------
coord <- data.frame(x=manhattan$LONGITUDE, y=manhattan$LATITUDE)
coordinates(coord) <- c("x", "y")
proj4string(coord) <- proj4string(manhattan)
coord <- spTransform(coord, CRS("+init=epsg:32618"))


## ----make a neighbors list----------------------------------------------------
nb <- poly2nb(manhattan, queen = T)


## ----figure 4-----------------------------------------------------------------
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(0.5, 0.5, 0.5, 0.5))
plot(manhattan, border = "white", col = "white", lwd=0.2)
plot(nb, cbind(manhattan$LONGITUDE, manhattan$LATITUDE), 
     col="darkgrey", add=TRUE, lwd=0.2, cex=0.1)
box()

```

*"Figure 4: Connections via poly2nb."* [Otani and Takahashi, 2001](https://www.jstatsoft.org/article/view/v099i13)

```{r, warning=FALSE, message=FALSE}
## ----run rflexscan------------------------------------------------------------
fls <- rflexscan(name = manhattan$DOHREGION,
  x = coord$x, y = coord$y, nb = nb,
  observed = manhattan$OBREAST, expected = manhattan$EBREAST)
summary(fls)
```  

^printed out chart of the underlying statistics calculated

```{r, warning=FALSE, message=FALSE}
## ----figure 6(a)--------------------------------------------------------------
par(mar = c(0.5, 0.5, 0.5, 0.5),
  oma = c(0.5, 0.5, 0.5, 0.5), lwd = 0.5, xaxt = 'n', yaxt = 'n')
plot(fls, rank = 1:7, edge.width = 0.5, col = brewer.pal(7, "RdYlBu"))
box(lwd = 1)
legend("topleft", legend = 1:7, fill = brewer.pal(7, "RdYlBu"), bty="n")


## ----figure 6(b)--------------------------------------------------------------
par(mar = c(0.5, 0.5, 0.5, 0.5),
    oma = c(0.5, 0.5, 0.5, 0.5), xaxt = 'n', yaxt = 'n')
choropleth(manhattan, fls, rank = 1:7, col = brewer.pal(7, "RdYlBu"))
legend("topleft", legend = 1:7, fill = brewer.pal(7, "RdYlBu"), bty="n")




```

*Figure 6: Significant clusters via the flexible scan statistic with the restricted likelihood ratio (p < 0.05).* [Otani and Takahashi, 2001](https://www.jstatsoft.org/article/view/v099i13)

#### **Spatial relative risk**  

+ Relative risk compares the risk of a health event (disease, injury, risk factor, or death) among one group with the risk among another group. [CDC](https://www.cdc.gov/csels/dsepd/ss1978/lesson3/section5.html#:~:text=A%20risk%20ratio%20(RR)%2C,the%20risk%20among%20another%20group)

+ Spatial relative risk is the ratio of densities describing spatial distribution of cases and controls. [Fernando and Hazelton, 2014](https://www.sciencedirect.com/science/article/abs/pii/S1877584513000555)

The following code is adapted from a  [tutorial](https://onlinelibrary.wiley.com/doi/10.1002/sim.7577) using the sparr R package and calculates RR based on KDE (same method used to find HR last week applied in a different context)

```{r, message=FALSE, warning=FALSE}
### Section 1 ###
library("sparr")

#-- Figure 1
data(pbc)
pbccas <- split(pbc)$case
pbccon <- split(pbc)$control

par(mfrow=c(1,2))
plot(pbccas,cex=.7,main="cases");box(bty="l");axis(1);axis(2)
title(ylab="Northing",xlab="Easting")
plot(pbccon,cex=.7,pch=5,main="controls");box(bty="l");axis(1);axis(2)
title(ylab="Northing",xlab="Easting")
```

*Figure 1: Primary biliary cirrhosis cases and sampled at-risk controls in a geographical region of northeast England* [Davies et al](https://onlinelibrary.wiley.com/doi/10.1002/sim.7577)  

```{r, warning=FALSE, message=FALSE}

#set common bandwith
hfix <- LSCV.risk(f=pbccas,g=pbccon, verbose = FALSE)

#compute fixed bandwidth estimate of PBC risk
f.tilde <- bivariate.density(pbccas,h0=hfix)
g.tilde <- bivariate.density(pbccon,h0=hfix)
rho.tilde <- risk(f=f.tilde,g=g.tilde)
pval.tilde <- tolerance(rs=rho.tilde,method="MC",ITER=200, verbose=FALSE)

# figure 5
plot(rho.tilde,xlab="Easting",ylab="Northing")
tol.contour(pim=pval.tilde,levels=c(0.01,0.05),lty=1:2,add=TRUE) #-- Figure 5 (right)
```

*Figure 5: Fixed-bandwidth spatial log-relative risk surfaces of the primary biliary cirrhosis (PBC) data estimated using sparr. Depicted is the recomputed estimate based on a jointly optimal, common case-control bandwidth found using LSCV.risk, as well as tolerance contours depicting significantly elevated risk of PBC calculated by Monte-Carlo simulations using tolerance. LSVC, least squares cross-validation* [Davies et al](https://onlinelibrary.wiley.com/doi/10.1002/sim.7577)  

```{r, warning=FALSE, message=FALSE}
#Adaptive bandwidth
hpilot.f <- BOOT.density(pbccas, verbose=FALSE)
hpilot.g <- BOOT.density(pbccon, verbose=FALSE)
h0 <- OS(pbc,nstar="geometric")
rho.hat1 <- risk(f=pbccas,g=pbccon,h0=h0,adapt=TRUE,hp=c(hpilot.f,hpilot.g),tolerate=TRUE)

plot(rho.hat1,zlim=c(-3.1,1.1),tol.args=list(levels=c(0.01,0.05),lty=1:2),xlab="Easting",ylab="Northing") #-- Figure 6 (left)
```

*Figure 6: Adaptive bandwidth spatial log-relative risk surfaces of the primary biliary cirrhosis (PBC) data estimated using sparr, withasymptotic (ASY) tolerance contours displayed. Method- an asymmetric estimate based on bootstrap-estimated pilot bandwidths and an over smoothing global bandwidth calculated using the pooled data.* [Davies et al](https://onlinelibrary.wiley.com/doi/10.1002/sim.7577)   

### Identifying spatial autocorrelation  
**Definitions:** 

+ Instead of a given variable (attribute) being distributed randomly across space, positive spatial autocorrelation indicates that spatially adjacent observations tend to have more similar data values [Haining 2001](https://www.sciencedirect.com/science/article/pii/B0080430767025110)

#### **Moran's I**  
Moran's I is a statistic that measures spatial autocorrelation. [More info](https://www.statisticshowto.com/morans-i/). 

[Tutorial source](https://mgimond.github.io/Spatial/spatial-autocorrelation-in-r.html)

```{r, warning=FALSE, message=FALSE}
library(tmap)
library(spdep)

load(url("https://github.com/mgimond/Spatial/raw/main/Data/moransI.RData"))

#visualize income distribution
map1<- tm_shape(s1) +
  tm_polygons(style="quantile", col ="Income") +
  tm_legend(outside = TRUE, text.size = .8) 

#define neigbors
nb <- poly2nb(s1, queen=TRUE)

#Assign weights to neigboring polygons
lw <- nb2listw(nb, style="W", zero.policy=TRUE)

#compute average neighbor income value for each polygon
Inc.lag <- lag.listw(lw, s1$Income)

#calculate moran's I statistic
I1 <- moran.test(s1$Income,lw) # analytic solution for p-val

#Monte Carlo sim value for p-value
I2 <- moran.mc(s1$Income, lw, nsim=599)


map1; I1; I2
```   

Interpretation: The Moran's I test statistic (0.283) is positive suggesting weak to moderate spatial clustering at a significant p-val (0.012-analytic,  0.027-MC).  


#### **Oden's Ipop**  
Oden's I is an adjustment to Moran's I that takes into account the underlying population density [Oden, 1995](https://pubmed.ncbi.nlm.nih.gov/7701154/).
-no real code for this online... need to look up the derivation and code manually. Will do when I have more time. 
